The mass dependence of dark matter halo alignments with large-scale structure

### 2017-07-15 draft.tex
Add files via upload

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{subfig}	% For figures position
\usepackage{epstopdf}	% To use .eps files with pdfLaTeX

@@

\newcommand*\dif{\mathop{}\!\mathrm{d}} 	% to have a proper differential symbol in integrals
\def\arraystretch{1.1}	% to control the spacing of columns

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Mass dependence of halo alignments]{The mass dependence of halo alignments with large-scale structure}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[D. Piras and more]{
Davide Piras$^{1}$\thanks{E-mail: davide.piras@studenti.unipd.it}
and more
% Benjamin Joachimi$^{2}$
\\
% List of institutions
$^{1}$Dipartimento di Fisica ``G. Galilei'', Universit\`{a} di Padova, via Marzolo 8, I-35131 Padova, Italy\\
%$^{2}$Department of Physics and Astronomy, University College London, Gower Street, London WC1E 6BT, UK
}

@@

% Abstract of the paper
\begin{abstract}
This is a simple template for authors to write new MNRAS papers.
The abstract should briefly describe the aims, methods, and main results of the paper.
It should be a single paragraph not more than 250 words (200 words for Letters).
No references should appear in the abstract.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
galaxies: halos --  dark matter -- large-scale structure of universe 
\end{keywords}

@@

\section{Introduction}
\label{sec:intro}
When looking at the sky, assuming a homogeneous and isotropic universe, one would expect to see randomly distributed galaxies with arbitrary orientation, whereas correlation between galaxy shapes is actually observed. Plenty of reasons can be exhibited to explain this, such as the effect of gravitational tidal fields from the surrounding dark matter haloes and the accretion of new material along favoured directions: this phenomenon is known as intrinsic alignment (sometimes referred to as IA; see \citet{TroxelIshak2015, Joachimietal2015, Kiesslingetal2015} for recent reviews). Besides carrying information about galaxy formation, IA contributes to the overall cosmic shear signal: \citet{Heavensetal2000} and \citet{CroftMetzler2000} first showed the non-negligible contamination of the weak lensing signal due to correlations in the intrinsic shapes of galaxies using simulated data, and many other works afterwards (\citet{Heymansetal2006, Sembolonietal2008}, just to mention two of them) confirmed that this effect must be taken into consideration in order not to bias the results from deep lensing surveys, such as LSST \citep{LSST2009} and Euclid \citep{Euclid2011}.

Measurements of the intrinsic alignment signal have been performed using data from hydrodynamical simulations \citep{Codisetal2015, Velliscigetal2015, Chisarietal2015, Hilbertetal2017, Tennetietal2016}, from N-body simulations \citep{Heavensetal2000, CroftMetzler2000, Jing2002, Kuhlenetal2007, Schneideretal2012}, as already mentioned, and considering observation data as well \citep{Mandelbaumetal2006, Hirataetal2007, Okumuraetal2009, Haoetal2011, vanUitertJoachimi2017}, claiming that massive red galaxies point towards matter overdensities, while blue galaxies do not show any clear hint of alignment \citep{Hirataetal2007, Mandelbaumetal2011}.

The dependence of the IA signal with redshift and mass has also been studied: \citet{Haoetal2011}, for example, studied a large sample of galaxy clusters from the SDSS DR7 finding a strong dependence of the alignment with redshift and an increasing trend with the mass of the brightest cluster galaxy. Moreover, \citet{Jing2002} used a $512^3$-particle simulation to get accurate formulae to fit the ellipticity correlation function over three orders of magnitude in mass, while an increasing trend of the IA amplitude with luminosity, or the corresponding mass, has been identified using real data, for galaxies \citep{Joachimietal2011, Singhetal2015} and for clusters \citep{vanUitertJoachimi2017}. In order to assume more stringent priors on the parameters of the models that will be adopted by future surveys, it is important to obtain accurate constraints on these dependences.

In this work, we delve into the dependence of the amplitude of the intrinsic alignment signal on the mass of the halo: after a brief theory overview (Sect.~\ref{sec:theory}), we analyse the trend for both galaxies and galaxy clusters using simulation and real data (Sect.~\ref{sec:data}), we perform a Bayesian analysis, described in Sect.~\ref{sec:method}, to establish the slope of the power-law model we assume to describe the IA amplitude signal, and show our results in Sect.~\ref{sec:resanddiscuss}.
\section{Theoretical background}
\label{sec:theory}
The physical picture of tidal interaction of a self-gravitating system that is in virial equilibrium with a velocity dispersion $\sigma$, such as an elliptical galaxy or a cluster of galaxies, would be a distortion of the system's gravitational potential through tidal gravitational forces. The particles of the system would remain in virial equilibrium and fill up the distorted potential along an isocontour of the gravitational potential, which would result in a change in the shape of the system. This shape should reflect the magnitude and the orientation of the tidal gravitational fields, and the magnitude of the change in shape should depend on how tightly the system is bound.

In isolated virialised systems the Jeans-equation would apply,
\begin{equation}
\frac{1}{\rho}\frac{\partial}{\partial r}\rho(\sigma^2) + \frac{2}{r}\beta\sigma^2 = 
-\frac{\partial\Phi}{\partial r},
\end{equation}
with the particle density $\rho$, distance $r$, gravitational potential $\Phi$ and the anisotropy parameter $\beta$, which we set to $\beta=0$ because we aim to derive only the scaling behaviour of the alignment amplitude. In this case, the Jeans-equation can be solved to yield $\rho\propto\exp(-\Phi/\sigma^2)$.

Gravitational tidal fields generated by the ambient large-scale structure would distort the gravitational potential, and we will work in the limit that the distortion is well-described by a second-order Taylor-expansion relative to the potential minimum,
\begin{equation}
\Phi\rightarrow\Phi+\frac{1}{2}\frac{\partial^2\Phi}{\partial r_i\partial r_j}r_ir_j
\end{equation}
with the distance $r_i$ from the minimum. Consequently, the density of particles would change according to
\begin{equation}
\rho\propto\exp\left(-\frac{\Phi}{\sigma^2}\right) \rightarrow 
\exp\left(-\frac{\Phi}{\sigma^2}\right)\left(1+\frac{1}{2\sigma^2}\frac{\partial^2\Phi}{\partial r_i\partial r_j}r_ir_j\right),
\end{equation}
and a measurement of the ellipticity through the second moments of $\rho$ would yield a proportionality to $R^2\partial^2\Phi/\sigma^2$, with $R$ the size of the object. The expansion is only valid in the limit of weak tidal fields, which is characterised by $R^2\partial^2\Phi/\sigma^2\ll 1$.

Now, the virial relationship $\sigma^2 = GM/R$ sets the velocity dispersion $\sigma^2$ into relation with the gravitational potential $-GM/R$, and with the scaling $M\propto R^3$ one would expect $\sigma^2\propto M^{2/3}$, such that $R^2/\sigma^2$ should be constant. Therefore, any scaling of the ellipticity with mass is entirely due to the dependence of tidal gravitational fields with the mass-scale.

The variance of tidal shear fields can be inferred from the variance of the matter density by the Poisson equation, $\Delta\Phi = 3\Omega_\mathrm{m}/(2\chi_H^2)\delta$, with the matter density $\Omega_m$ and the Hubble-distance $\chi_H=c/H_0$. Computing the tidal shear fields $\partial^2\Phi$ shows that they must have the same fluctuation statistics as the density field $\delta$. In Fourier-space, the solution to the Poisson equation is $\Phi \propto \delta/k^2$ with the wave vector $k$, and the tidal shear fields would become $k_ik_j\Phi \propto k_ik_j/k^2\:\delta$. Therefore, the spectrum $P_{\partial^2\Phi}(k)$ of the tidal shear fields is proportional to the spectrum $P_\delta(k)$ of the density fluctuations.

Consequently, one can derive the variance of tidal shear fields from the variance of the density fluctuations, i.e.\ from the CDM-spectrum $P_\delta(k)$. For doing that, one can relate a mass scale $M$ to the wave vector $k$ by requiring that the mass $M$ should be contained in a sphere of radius $R$, $M=4\pi\Omega_\mathrm{m}\rho_\mathrm{crit}R^3/3$, which in turn defines a scale $k$ in the spectrum which is proportional to $M^{-1/3}$. This implies that on galaxy- and cluster-scales, where the CDM-spectrum scales $\propto k^{-2\ldots-3}$, one obtains for the standard deviation of the tidal shear field a behaviour $\propto M^{1/2\ldots 1/3}$. We compare these very general theoretical predictions with our results in Sect.~\ref{sec:resanddiscuss}.
\section{Data}
\label{sec:data}
\subsection{Simulations}
\label{subsec:sim}
In this work we consider haloes from two different simulations: 
\begin{enumerate}
\item the \textbf{Millennium Simulation (MS)}, first presented in \citet{Springeletal2005}, which uses $2160^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{MS}} =1.2 \times 10^9 M_{\sun}$ enclosed in a 500 Mpc$/h$-side box to sample dark matter haloes and study their growth. In particular, we take into consideration 2 of its 64 snapshots, i.e.\ the one at $z = 0$ (snapshot $63$), which we use for all our analysis, and the one at $z \simeq 0.46$ (snapshot $49$), which is studied only as an aside. Dark matter haloes are identified as in \citet{Joachimietal2013a} and references therein, assuming that all galaxies share the three-dimensional shape of their host halo. All haloes have a minimum number of particles $N_{\mathrm{P}} = 300$.
\item The \textbf{Millennium-XXL Simulation (XXL)}, which samples $6720^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{XXL}} = 8.456 \times 10^9 M_{\sun}$ confined in a cubic region of $3000$ Mpc$/h$ on a side \citep{Anguloetal2012}; in this case, we consider only one snapshot, at $z = 0$. Haloes are selected using an ellipsoidal overdensity algorithm, as described in \citet{Despalietal2013} and \citet{Bonamigoetal2015}, assuming a minimum number of particles $N_{\mathrm{P}} = 1000$.
\end{enumerate} 
%the shape of the haloes is then derived as for the MS from the eigenvectors and eigenvalues of the mass tensor 
%\begin{equation}
%    \mathbf{M}_{\mu \nu} = m_P \sum_{i=1}^{N_P} \frac{r_{i, \mu} r_{i, \nu}}{M_{TOT}} = \frac{1}{N_P} \sum_{i=1}^{N_P} r_{i, \mu} r_{i, \nu}
%	\label{eq:mt}
%\end{equation}
%with $M_{TOT}$ the total mass of the object. Note that the mass tensor (Eq.~\ref{eq:mt}) and the inertia tensor (Eq.~\ref{eq:sit}) are different quantities \citep{Bettetal2007}, but the analysis on the two simulations yields pretty consistent results, as will be shown; also, for this catalogue no ``reduced'' tensor is taken into consideration.

In the simulations the same set of cosmological parameters is adopted, namely they both assume a spatially flat $\Lambda$CDM universe with the total matter density $\Omega_{\mathrm{m}} = \Omega_{\mathrm{b}} + \Omega_{\mathrm{dm}} = 0.25$, where $\Omega _{\mathrm{b}} = 0.045$ indicates baryons and $\Omega_{\mathrm{dm}} = 0.205$ represents dark matter, a cosmological constant $\Omega_{\Lambda} = 1 - \Omega_{\mathrm{m}} = 0.75$, the Hubble parameter $h = 0.73$ and the density variance in spheres of radius $8 \ \mbox{Mpc}/h$ $\sigma_8 = 0.9$. All the density parameters are in units of the critical density.

To compare our results with real data, we also take into consideration the analysis presented in \citet{vanUitertJoachimi2017}: in that work, the clusters contained in the redMaPPer catalogue \citep{Rykoffetal2014} version 6.3 were used to constrain the intrinsic alignment signal amplitude $A_{\mathrm{IA}}$, which was then studied as a function of the halo mass.

\subsection{Halo shapes}
\label{subsec:haloshapes}
We define the simple inertia tensor\footnote{MS and XXL use two different tensors to describe the shape, but our choice has no impact on the ellipticity of the haloes (see also \citet{Bettetal2007} for further details).}, whose eigenvalues and eigenvectors describe the shape of the halo, as:
\begin{equation}
    \mathbf{M}_{\mu \nu} \propto \sum_{i=1}^{N_{\mathrm{P}}} r_{i, \mu} r_{i, \nu} \ ,
	\label{eq:sit}
\end{equation}
where $N_{\mathrm{P}}$ is the total number of particles within the halo, and $\mathbf{r}_{i}$ is the vector that indicates the position of the $i-$th particle with respect to the centre of the halo, i.e.\ the location of the gravitational potential minimum. For the MS only, we also consider a reduced inertia tensor, which is defined as \citep{Pereiraetal2008}:
\begin{equation}
    \mathbf{M}_{\mu \nu}^{\mathrm{red}} \propto \sum _{i=1}^{N_{\mathrm{P}}} \frac{r_{i, \mu} r_{i, \nu}}{r_i^2} \ ,
	\label{eq:rit}
\end{equation}
with $r_i^2$ the square of the three-dimensional distance of the i-th particle from the centre of the halo; the reduced inertia tensor is more weighted towards the centre of the halo, and may yield a more reliable approximation of the shape of the galaxy \citep{Joachimietal2013b, Chisarietal2015}.

The eigenvectors define an ellipsoid, which we project onto one of the faces of the cube that hosts the simulation: the resulting ellipse is the shape of the galaxy, which is assumed to be the same as the shape of the host halo. We proceed as in \citet{Joachimietal2013a, Joachimietal2013b} to define the ellipticity $\epsilon$ of the galaxy, considering our objects as early-type galaxies: indicating the three unit eigenvectors as $\mathbf{s}_{\mu} = \big \{ s_{x, \mu}, s_{y, \mu}, s_{\mathbin{\|}, \mu}\big \}^\intercal$ and the absolute values of the semi-axes as $\omega_{\mu}$, $\mu \in \{1,2,3\}$, we define a symmetric tensor
\begin{equation}
    \mathbf{W}^{-1} = \sum _{\mu=1}^{3} \frac{s_{\perp, \mu} s^\intercal_{\perp, \mu}}{\omega_{\mu}^2} - \frac{\mathbf{\kappa}\mathbf{\kappa}^{\intercal}}{\alpha^2} \ ,
	\label{eq:symtensor}
\end{equation}
with $s_{\perp, \mu} = \big \{ s_{x, \mu}, s_{y, \mu}\big \}^\intercal$ the eigenvector projected along the line of sight, 
\begin{equation}
   \mathbf{\kappa} = \sum_{\mu = 1}^{3} \frac{s_{\mathbin{\|}, \mu} \mathbf{s}_{\perp, \mu}}{\omega_{\mu}^2} \ ,
	\label{eq:kappa}
\end{equation}
and 
\begin{equation}
    \alpha^2=\sum_{\mu = 1}^{3} \bigg( \frac{s_{\mathbin{\|}, \mu}}{\omega_{\mu}} \bigg)^2 \ .
	\label{eq:alpha}
\end{equation}
We then compute the two components of the ellipticity \citep{BartelmannSchneider2001}:
\begin{align}
    	\epsilon_{1} = \frac{W_{11} - W_{22}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}}\ ; \\ 
           \epsilon_{2} = \frac{2 W_{12}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}} \ .
\end{align}

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{subfig}	% For figures position
\usepackage{epstopdf}	% To use .eps files with pdfLaTeX
\usepackage{multirow}	% To have multirows in columns

@@

\newcommand*\dif{\mathop{}\!\mathrm{d}} 	% to have a proper differential symbol in integrals
%\def\arraystretch{1.1}	% to control the spacing of columns
%\newcommand{\appropto}{\mathrel{\vcenter{
  %\offinterlineskip\halign{\hfil$##$\cr
    %\propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}} % to have a very ugly symbol of similar-propto
\newenvironment{correction}
 {\bfseries}% \begin{boldenv}
  {}% \end{boldenv}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[The mass dependence of dark matter halo alignments]{The mass dependence of dark matter halo alignments with large-scale structure}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[D. Piras et al.]{
Davide Piras$^{1, 2}$\thanks{E-mail: d.piras@ucl.ac.uk; davide.piras@studenti.unipd.it},
Benjamin Joachimi$^{1}$,
Bj\"{o}rn Malte Sch\"{a}fer$^{3}$,
Mario Bonamigo$^{4,5}$, 
\newauthor{Stefan Hilbert$^{6,7}$,
and Edo van Uitert$^{1}$}
\\
% List of institutions
$^{1}$Department of Physics and Astronomy, University College London, Gower Street, London WC1E 6BT, UK\\
$^{2}$Dipartimento di Fisica ``G. Galilei'', Universit\`{a} di Padova, via Marzolo 8, I-35131 Padova, Italy\\
$^{3}$Astronomisches Recheninstitut, Zentrum f\"{u}r Astronomie der Universit\"{a}t Heidelberg, Philosophenweg 12, 69120 Heidelberg, Germany\\
$^{4}$Dark Cosmology Centre, Niels Bohr Institute, University of Copenhagen, Juliane Maries Vej 30, DK-2100 Copenhagen, Denmark\\
$^{5}$Aix Marseille Universit\'e, CNRS, LAM (Laboratoire d'Astrophysique de Marseille) UMR 7326, 13388, Marseille, France\\
$^{6}$Exzellenzcluster Universe, Boltzmannstr. 2, 85748 Garching, Germany\\
$^{7}$Ludwig-Maximilians-Universit{\"a}t, Universit{\"a}ts-Sternwarte, Scheinerstr. 1, 81679 M{\"u}nchen, Germany
}

@@

% Abstract of the paper
\begin{abstract}
Tidal gravitational forces can modify the shape of galaxies and clusters of galaxies, thus correlating their orientation with the surrounding density field. We study the dependence of this phenomenon, known as intrinsic alignment (IA), on the mass of the dark matter haloes that host these bright structures, analysing the Millennium and Millennium-XXL $N$-body simulations. We closely follow the observational approach, measuring the halo position-halo shape alignment and subsequently dividing out the dependence on halo bias.
%To quantify the intrinsic alignment signal, we adopt an estimator that mimics the observational approach, and divide it by the clustering signal to dispose of the cluster bias factor.
%We compare our theoretical predictions for the scaling of the IA with mass in a dark matter universe with the results of our Bayesian analysis, assuming for the intrinsic alignment amplitude a power-law model with slope $\beta_{\mathrm{M}}$. 
We derive a theoretical scaling of the IA amplitude with mass in a dark matter universe, and predict a power-law with slope $\beta_{\mathrm{M}}$ in the range $1/3$ to $1/2$, depending on mass scale.
% and test it on simulation samples through Bayesian analysis.
% assuming for the intrinsic alignment amplitude a power-law model with slope $\beta_{\mathrm{M}}$. 
We find that the simulation data agree with each other and with the theoretical prediction remarkably well over three orders of magnitude in mass, with the joint analysis yielding an estimate of $\beta_{\mathrm{M}} = 0.36^{+0.01}_{-0.01}$. This result does not depend on redshift or on the details of the halo shape measurement. The analysis is repeated on observational data, obtaining a significantly higher value, $\beta_{\mathrm{M}} = 0.56^{+0.05}_{-0.05}$.\begin{correction} There are also small but significant deviations from our simple model in the simulation signals at both the high- and low-mass end. We discuss possible reasons for these discrepancies and argue that they can be attributed to physical processes not captured in the model or the dark matter-only simulations.\end{correction}
%and claim that using data from hydrodynamical (instead of $N$-body) simulations could lead to a value of $\beta_{\mathrm{M}}$ compatible with real data.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
galaxies: haloes -- dark matter -- large-scale structure of Universe 
\end{keywords}

@@

\section{Introduction}
\label{sec:intro}
%In a statistically isotropic universe, one would expect galaxy images not to have preferred orientations; however, we observe that galaxy shapes are locally correlated with the surrounding large-scale structure, and therefore with each other. 
In a statistically isotropic universe, galaxy images have no globally preferred orientation; this, however, does not preclude local correlations of galaxy shapes, with each other and with the large-scale structure.
Several mechanisms have been proposed to explain this, such as the accretion of new material along favoured directions, and the effect of gravitational tidal fields from the surrounding dark matter distribution. In this latter picture, in particular, the shape of luminous structures is affected by tidal interactions by the surrounding dark matter halo, and, consequently, galaxies and clusters also align \citep{Kiesslingetal2015}.

This phenomenon is known as intrinsic alignment (henceforth IA; see \citealt{TroxelIshak2015, Joachimietal2015, Kiesslingetal2015, Kirketal2015} for recent reviews): besides carrying information about galaxy formation, intrinsic alignment has an impact on the measurements of the cosmological weak lensing effect, the induced correlations of galaxy image shapes due to gravitational lensing by the mass distribution along the line of sight. IA must therefore be taken into consideration when analysing lensing surveys, such as LSST \citep{LSST2009} and Euclid \citep{Euclid2011}. \citet{Heavensetal2000} and \citet{CroftMetzler2000} first predicted the non-negligible contamination of the weak lensing signal due to correlations in the intrinsic shapes of galaxies, and a number of works afterwards (e.g.\ \citealt{Heymansetal2006, Sembolonietal2008, Kirketal2012}) confirmed that this effect must be accounted for in order not to bias the cosmological inference.
%the results of the observations.

The intrinsic alignment signal has been measured in $N$-body simulations \citep{Heymansetal2006, Kuhlenetal2007, Leeetal2008, Schneideretal2012, Joachimietal2013a, Joachimietal2013b}, in hydrodynamical simulations \citep{Codisetal2015, Velliscigetal2015b, Chisarietal2015, Tennetietal2016, Hilbertetal2017}, and in imaging data \citep{Joachimietal2011, Haoetal2011, Lietal2013, Singhetal2015, vanUitertJoachimi2017}. Results from hydrodynamical simulations and observations, in particular, claimed that massive red galaxies point towards matter overdensities, while blue galaxies do not show any clear sign of alignment \citep{Hirataetal2007, Mandelbaumetal2011}.

The amplitude of the IA signal has been found to increase with mass in observations. \citet{Haoetal2011}, for example, studied a large sample of galaxy clusters from the SDSS DR7 finding a dependence of the alignment with the mass of the brightest cluster galaxy, and an increasing IA amplitude with luminosity, or the corresponding mass, has been identified for luminous red galaxies (LRGs, \citealt{Joachimietal2011, Singhetal2015}) and for galaxy clusters \citep{vanUitertJoachimi2017}.

Simulations agree with this picture: using a $512^3$-particle $N$-body simulation, \citet{Jing2002}, for example, found that the alignment increases with mass over three orders of magnitude, up to about $10^{13} M_{\sun}$. Moreover, \citet{Leeetal2008} used data from the Millennium simulation, and claimed stronger correlations with higher mass over two mass bins around $10^{12} M_{\sun}$, while \citet{Joachimietal2013b}, using the same set of data, found an increasing trend over the same mass range only for early-type galaxies (whose shapes were assumed to follow those of the underlying haloes), while no dependence on luminosity or particular trend with mass or luminosity for late-type galaxies (whose orientation was determined by the halo spin) was identified.

These results motivate us to search for a universal relation between the alignment strength and the mass of dark matter haloes. This could lend support to an IA mechanism that successfully explains the trends; moreover, we need to reduce the degrees of freedom in modelling the IA signal to obtain tighter cosmology constraints from lensing observations. To achieve this, we delve into the dependence of the amplitude of the intrinsic alignment signal on the mass of the halo, which is postulated to be the main driver of the shape and orientation of the hosted galaxies (e.g.\ \citealt{Joachimietal2015}). We first derive the expected scaling from the theory, and then test our predictions using data from two $N$-body simulations, mimicking the observational approach in order to be able to straightforwardly compare our results with real data.

We give details about our theoretical model and derive the expected scaling of the IA with halo mass in the tidal alignment paradigm in Sect.~\ref{sec:theory}. We then present the simulations that we use (Sect.~\ref{subsec:sim}), and how we define the shapes of the dark matter haloes they contain (Sect.~\ref{subsec:haloshapes}). In Sect.~\ref{subsec:measurements} we explain how we measure the intrinsic alignment signal, and then (Sect.~\ref{subsec:modelling}) we describe our mass-dependent IA model. We finally show our results and compare them with our theoretical predictions (Sect.~\ref{subsec:simdatard}) and real data (Sect.~\ref{subsec:obsdatard}).

\section{Theoretical background}
\label{sec:theory}
The physical picture of tidal interaction of a self-gravitating system that is in virial equilibrium, with a velocity dispersion $\sigma^2$, such as \begin{correction}an elliptical or a relaxed galaxy cluster\end{correction}, is a distortion of the system's gravitational potential through tidal gravitational forces. The particles of the system remain in virial equilibrium and fill up the distorted potential along an isocontour of the gravitational potential, which results in a change in the shape of the system. This shape modification reflects the magnitude and the orientation of the tidal gravitational fields, and the magnitude of the change in shape depends on how tightly the system is bound.

In isolated virialised systems the Jeans-equation applies \citep[see e.g.][]{BinneyTremaine2008}:
\begin{equation}
\frac{1}{\rho}\frac{\partial}{\partial r}(\rho\sigma^2) + \frac{2}{r}\beta_\mathrm{aniso}\sigma^2 = 
-\frac{\partial\Phi}{\partial r} \ ,
\end{equation}
with $\rho$ the particle density, $r$ the distance, $\Phi$ the gravitational potential and $\beta_\mathrm{aniso}$ the anisotropy parameter, which we set to $\beta_\mathrm{aniso}=0$ because we aim to derive only the scaling behaviour of the alignment amplitude. In the case of vanishing anisotropy and a constant velocity dispersion, the Jeans-equation can be solved to yield an exponential dependence $\rho\propto\exp(-\Phi/\sigma^2)$ between the density and the gravitational potential.

\begin{correction}
Gravitational tidal fields generated by the ambient large-scale structure distort the gravitational potential $\Phi$, and we will work in the limit that the distortion is well described by a second-order Taylor-expansion of the potential relative to the centre of the galaxy at $\bmath{r}_0 = 0$. Therefore, the potential relevant for the motion of test particles is given by
\begin{equation}
\Phi(\bmath{r}) \rightarrow
\Phi(\bmath{r}) + \frac{1}{2}\sum_{i,j=1}^{3}\frac{\partial^2\Phi(\bmath{r}_0)}{\partial r_i\partial r_j}r_ir_j \ ,
\end{equation}
with $i,j$ indicating the spatial dimensions. Consequently, the density of particles changes according to
\begin{equation}
\label{eq:density}
\rho = \rho_0
\left(1-\frac{1}{2\sigma^2}\sum_{i,j=1}^{3}\frac{\partial^2\Phi(\bmath{r}_0)}{\partial r_i\partial r_j}r_ir_j\right) \ ,
\end{equation}
with $\rho_0 \propto \exp\left(-\frac{\Phi(\bmath{r})}{\sigma^2}\right)$, under the assumption of a weak distortion such that a Taylor-expansion of the exponential to first order is sufficient.

The projected ellipticity is calculated through the tensor of second brightness moments \citep{BartelmannSchneider2001}, which we define using the density $\rho$:
\begin{equation}
q_{ij} = \int \dif^2x \ \rho(\bmath{x}) x_i x_j \ ,
\end{equation}
where the integral is calculated over the plane of the projected sky, with coordinates $x_i,x_j$ and $i,j \in \{1,2\}$. If we consider the distortion of the density described in Eq.~(\ref{eq:density}), we obtain:
\begin{multline}
\label{eq:bridis}
\widetilde{q_{ij}} = \int \dif^2x \ \rho_0(\bmath{x}) \left(1-\sum_{a,b=1}^2\frac{\partial_{a,b}^2\Phi(\bmath{x}_0)}{2\sigma^2} x_a x_b\right) x_i x_j \simeq \\ 
\simeq \int \dif^2 x \ \rho_0(\bmath{x}) x_i x_j - \sum_{a,b=1}^2\frac{\partial_{a,b}^2\Phi}{2\sigma^2} \int \dif^2 x \ \rho_0(\bmath{x}) x_a x_b x_i x_j = \\ = q_{ij} + \psi_{ij} \ ,
\end{multline}
where the equivalence is not exact due to the approximation that $\partial_{a,b}^2\Phi(\bmath{x}_0) / \sigma^2$ is constant and can be thus taken out of the integral. From this latter equation, we can see that because of the tidal shear fields the second moments of the brightness distribution get a correction $\psi_{ij}$: this term is small in the limit of weak tidal fields, which is characterised by $R^2\partial^2\Phi/\sigma^2\ll 1$ with $R$ being the size of the halo and providing a bound for $r$.

If we then define the complex ellipticity as in \citet{BartelmannSchneider2001}:
\begin{equation}
\label{eq:ell}
\varepsilon = \frac{q_{11} - q_{22}}{q_{11} + q_{22}+Q} + 2\mathrm{i}\frac{q_{12}}{q_{11} + q_{22}+Q} \ ,
\end{equation} 
with $Q=2\sqrt{q_{11}q_{22}-q_{12}^2}$, and consider the correction obtained in Eq.~(\ref{eq:bridis}), we can write:
\begin{equation}
\label{eq:epsilon}
\widetilde{\varepsilon} = \varepsilon +  \psi_{\epsilon}\ ,
\end{equation} 
with $\varepsilon$ as in Eq.~(\ref{eq:ell}), and
\begin{equation}
\psi_{\epsilon} = \frac{\psi_{11} - \psi_{22}}{q_{11} + q_{22}+Q} + 2i\frac{\psi_{12}}{q_{11} + q_{22}+Q} \ ,
\end{equation}
assuming that $\psi_{ij}$ is a small correction.
\end{correction}

%Gravitational tidal fields generated by the ambient large-scale structure distort the gravitational potential $\Phi$, and we work in the limit that the distortion is well-described by a second-order Taylor-expansion relative to the potential minimum at $\bmath{r}_0$,
%\begin{equation}
%\Phi(\bmath{r}) = \Phi(\bmath{r}_0) + \frac{1}{2}\sum_{i,j=1}^{3}\frac{\partial^2\Phi(\bmath{r}_0)}{\partial r_i\partial r_j}(\bmath{r}-\bmath{r}_0)_i(\bmath{r}-\bmath{r}_0)_j \ ,
%\end{equation}
%with $i,j$ indicating the spatial dimensions. Consequently, the density of particles would change according to
%\begin{equation}
%\rho \propto 
%\exp\left(-\frac{\Phi(\bmath{r}_0)}{\sigma^2}\right)
%\times
%\left(1-\frac{1}{2\sigma^2}\sum_{i,j=1}^{3}\frac{\partial^2\Phi(\bmath{r}_0)}{\partial r_i\partial r_j}(\bmath{r}-\bmath{r}_0)_i(\bmath{r}-\bmath{r}_0)_j\right) \ ,
%\end{equation}
%under the assumption of a weak distortion such that a Taylor-expansion of the exponential to first order is sufficient. Physically, this would be the limit of weak tidal fields, which is characterised by $R^2\partial^2\Phi/\sigma^2\ll 1$ with $R$ being the size of the halo, which provides a bound for $\left|\bmath{r}-\bmath{r}_0\right|$.

Therefore, a measurement of the ellipticity of a tidally distorted halo through the second moments of the projected density yields a proportionality to $R^2\partial^2\Phi/\sigma^2$, in accordance with the linear alignment model for elliptical galaxies, where the tidally induced ellipticity is proportional to the magnitude of the tidal fields \citep{HirataSeljak2004, Blazeketal2011, Blazeketal2015}. It is worth pointing out that the tidal effect on galaxies is determined by the velocity scale $\sigma^2$, whereas in the case of gravitational lensing the tidal shear is measured in units of the velocity scale $c^2$.

In the case of virialised systems it is possible to relate the velocity dispersion $\sigma^2$ with the size of the object $R$: the virial relationship $\sigma^2 = GM/R$ assumes a proportionality between the specific kinetic energy, $\sigma^2/2$, and the magnitude of the specific potential energy, $GM/R$. With the scaling $M\propto R^3$ one expects $\sigma^2\propto M^{2/3}$, such that $R^2/\sigma^2$ is constant. Therefore, any scaling of the ellipticity with mass is entirely due to the dependence of tidal gravitational fields with the mass-scale, and more massive systems are subjected to stronger tidal interactions because of the stronger fluctuations in the tidal fields that they are experiencing.

The variance of tidal shear fields can be inferred from the variance of the matter density by the Poisson equation, $\Delta\Phi = 3\Omega_\mathrm{m}/(2\chi_H^2)\delta$, with $\Omega_\mathrm{m}$ the matter density, $\chi_H=c/H_0$ the Hubble-distance, and $\delta$ the density field. Computing the tidal shear fields $\partial^2\Phi$ shows that they must have the same fluctuation statistics as $\delta$: in Fourier-space, the solution to the Poisson equation is $\Phi \propto \delta/k^2$ with the wave vector $k$, and the tidal shear fields become $k_ik_j\Phi \propto k_ik_j/k^2\:\delta$. Therefore, the power spectrum $P_{\partial^2\Phi}(k)$ of the tidal shear fields is proportional to the power spectrum $P_\delta(k)$ of the density fluctuations.
